---
title: "Comparison of Performance - Alexandrov et al. data"
author: "Daan Hazelaar"
date: '2022-07-08'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

```{r}
load(file = "../data/resultsAlexandrovAllTools.RData")
load(file = "../data/alexandrov_data_processed.RData")
load(file = "../data/sampleOrigin.RData")
source(file = 'function_calculateMetrics.R')
source(file = "function_calculateConfusionVariants.R")

theme_Job <- ggplot2::theme(
    legend.position = 'bottom',
    legend.direction = 'horizontal',
    text = ggplot2::element_text(size=9, family='Helvetica', face = 'bold'),
    axis.text.x = ggtext::element_markdown(),
    axis.title.x = ggtext::element_textbox_simple(width = NULL, halign = .5),
    axis.title.y = ggtext::element_textbox_simple(size = 8, orientation = 'left-rotated', width = NULL, halign = .5),
    strip.text = ggtext::element_textbox_simple(width = NULL, halign = .5),
    panel.grid.major.x = ggplot2::element_line(colour = 'grey90', linetype = 'dotted'),
    panel.grid.major.y = ggplot2::element_line(colour = '#E5E5E5', linetype = 'dotted'),
    panel.grid.minor.y = ggplot2::element_blank(),
    panel.background = ggplot2::element_rect(fill = NA, colour = 'black'),
    panel.border = ggplot2::element_rect(fill = NA, colour = NA),
    strip.background = ggplot2::element_rect(colour = 'black', fill = 'white'),
    legend.text = ggtext::element_markdown()
)
```


Descriptive statistic of the WGS samples in the alexandrov dataset
```{r}
fociPerSample <- alexandrovData$reportedKataegis |> 
    tibble::as_tibble() |> 
    dplyr::group_by(.data$sampleNames) |> 
    dplyr::summarise(
        nKataegisFoci = dplyr::n()
    )

alexandrovData$genomicVariants |>
    tibble::as_tibble() |> 
    dplyr::group_by(.data$sampleNames) |> 
    dplyr::summarise(
        .groups = "keep",
        totalVariants = dplyr::n(),
        kataegisVariants = sum(.data$kataegis),
        tissue = unique(.data$tissue)
    ) |> 
    dplyr::mutate(
        TMB = base::round(.data$totalVariants / 3095, 3),
        TMBcat = dplyr::case_when(
            TMB < 1 ~ "low",
            TMB >= 1 & TMB < 10 ~ "mid",
            TMB >= 10 ~ "high"
        )
    ) |> 
    dplyr::left_join(fociPerSample, by = "sampleNames") |> 
    dplyr::mutate(nKataegisFoci = tidyr::replace_na(.data$nKataegisFoci, 0)) |> 
    dplyr::group_by(TMBcat) |> 
    dplyr::summarise(
        nSample = dplyr::n(),
        nKataegisSamples = base::sum(.data$nKataegisFoci != 0),
        totalVariants = base::sum(.data$totalVariants),
        kataegisVariants = base::sum(.data$kataegisVariants),
        nKataegisFoci = base::sum(.data$nKataegisFoci)
    )
```

Combine results into a dataframe and add columns needed for further analysis.
```{r}
kataegisSampleNames <- base::unique(alexandrovData$reportedKataegisFoci$sampleNames)
nVariantsInSample <- base::lapply(alexandrovData$genomicVariants, base::length)
nVariantsInSampleTib <- tibble::tibble(sampleNames = base::names(nVariantsInSample), totalVariantsInSample = base::unlist(nVariantsInSample))

# combine all results in a single tibble and add columns needed for subsequent analysis
resultsAlex <- dplyr::bind_rows( 
    "MafTools" = resultsAlexandrovAllTools$maftools, 
    "SeqKat" = resultsAlexandrovAllTools$seqkat,
    "ClusteredMutations" = resultsAlexandrovAllTools$clusteredMutations,
    "katdetectr" = resultsAlexandrovAllTools$katdetectr,
    "kataegis" = resultsAlexandrovAllTools$kataegis,
    "SigProfilerClusters" = resultsAlexandrovAllTools$sigProfilerClusters,
    .id = "package") |> 
    dplyr::distinct() |> 
    dplyr::left_join(nVariantsInSampleTib, by = "sampleNames") |> 
    dplyr::mutate(
        # calculate tumor mutational burden per sample by total number of variants / (length of the genome / 10^6) ~ 3095
        TMB = base::round(totalVariantsInSample / 3095, 3),
        TMBcat = dplyr::case_when(
            TMB < 1 ~ "low",
            TMB >= 1 & TMB < 10 ~ "mid",
            TMB >= 10 ~ "high"
        ),
        TMBcat = factor(TMBcat, levels = c("high", "mid", "low")),
        # specify if a kataegis foci was detected by the method
        detectedKataegisInSample = dplyr::if_else(is.na(.data$totalVariants), FALSE, TRUE),
        # specify if a sample contains a kataegis foci according to Alexandrov
        kataegisInSample = dplyr::if_else(sampleNames %in% kataegisSampleNames, TRUE, FALSE)
    ) |> 
    dplyr::left_join(sampleOrigin, by = "sampleNames")
```

Calculate the performance of the kataegis detection tools regarding sample classfication
```{r}
resultsPerSampleAlex <- resultsAlex |> 
    # summarise if a method detected kataegis in a sample
    dplyr::group_by(.data$package, .data$sampleNames, .data$TMB, .data$TMBcat, .data$runTime) |> 
    dplyr::summarise(
        .groups = "keep",
        # is a kataegis foci detected by the method
        detectedKataegisInSample = any(detectedKataegisInSample),
        # is there actualy a kataegis foci in that sample
        kataegisInSample = any(kataegisInSample)
    ) |> 
    dplyr::ungroup() |> 
    dplyr::rowwise() |> 
    # calculate confusion matrix
    dplyr::mutate(
        TP = .data$detectedKataegisInSample == TRUE & .data$kataegisInSample == TRUE,
        FP = .data$detectedKataegisInSample == TRUE & .data$kataegisInSample == FALSE,
        TN = .data$detectedKataegisInSample == FALSE & .data$kataegisInSample == FALSE,
        FN = .data$detectedKataegisInSample == FALSE & .data$kataegisInSample == TRUE
    ) |> 
    dplyr::ungroup()

performancePerSampleAlex <- resultsPerSampleAlex |> 
    dplyr::group_by(.data$package) |> 
    dplyr::group_modify(~ calculateMetrics(.x), .keep = TRUE) |> 
    dplyr::ungroup()

performancePerSampleTMB <- resultsPerSampleAlex |> 
    dplyr::group_by(.data$package, .data$TMBcat) |> 
    dplyr::group_modify(~ calculateMetrics(.x), .keep = TRUE) |> 
    dplyr::ungroup()
```

Calculate the performance of the kataegis detection tools regarding variant classification
```{r}
resultsPerVariant <- resultsAlex |>
    dplyr::group_by(.data$package, .data$sampleNames) |> 
    dplyr::group_modify(~ calculateConfusionVariants(.x, variantData = alexandrovData), .keep = TRUE) |> 
    dplyr::ungroup()

performancePerVariantAlex <- resultsPerVariant |> 
    dplyr::group_by(.data$package) |> 
    dplyr::group_modify(~ calculateMetrics(.x), .keep = TRUE) |> 
    dplyr::ungroup() |> 
    arrange(desc(nMCC)) |> 
    dplyr::select(package, accuracy, nMCC, F1, TPR, TNR)

performancePerVariantTMBAlex <- resultsPerVariant |> 
    dplyr::group_by(.data$package, .data$TMBcat) |> 
    dplyr::group_modify(~ calculateMetrics(.x), .keep = TRUE) |> 
    dplyr::ungroup()
```

Plot nMCC as a function of the TMB for each package
```{r}
levelOrderTMB <- c("Low (TMB < 0.1)", "Mid (0.1 >= TMB < 10)", "High (TMB >= 10)")
levelOrderTMB <- c("low", "mid", "high")
levelOrderPackages <- factor(performancePerVariantAlex$package, levels = arrange(performancePerVariantAlex, nMCC)$package)

colors = c('ClusteredMutations' = '#66C2A5', 'kataegis' = '#FC8D62', 'katdetectr' = '#8DA0CB', 'MafTools' = '#E78AC3', 'SeqKat' = '#A6D854', 'SigProfilerClusters' = '#FFD92F')

plotnMMC.Alex <- performancePerVariantTMBAlex |> 
    ggplot2::ggplot(aes(
        x = factor(TMBcat, levels = levelOrderTMB), 
        y = nMCC, 
        group = factor(package, levels = levelOrderPackages), 
        colour = factor(package, levels = levelOrderPackages))) +
    ggplot2::geom_line(size = 0.5, alpha = 0.4, lty = 11) +
    ggplot2::geom_point(aes(x = factor(TMBcat, levels = levelOrderTMB), y = nMCC, fill = package), col = 'black', shape = 21, lwd = 2, position = ggbeeswarm::position_quasirandom(width = .05)) +
    ggplot2::scale_fill_manual(values = colors) +
    ggplot2::scale_color_manual(values = colors, guide = 'none') +
    ggplot2::scale_x_discrete(labels=c(
        "low" = "Low (TMB < 0.1)<br>n = 301", 
        "mid" = "Mid (0.1 ≥ TMB < 10)<br>n = 186",
        "high" = "High (TMB ≥ 10)<br>n = 20")
    ) +
    ggplot2::ylim(0.5, 1) +
    ggplot2::labs(x = "TMB class", y = 'nMCC<br>Alexandrov et al.') +
    theme_Job
```



plot performance in variant classification split on TMB
```{r}
# facet labels
TMB.labs <- c("TMB >= 10", "0.1 >= TMB < 10", "TMB < 0.1")
names(TMB.labs) <- c("high", "mid", "low")

plotPerformancePerVariantTMB <- performancePerVariantTMBAlex |> 
    tidyr::pivot_longer(
        cols = c(accuracy, TPR, TNR, F1, nMCC),
        names_to = "performanceMetric",
        values_to = "value"
    ) |> 
    ggplot2::ggplot(aes(y = factor(package, levels = arrange(performancePerVariantAlex, nMCC)$package), 
                        x = factor(performanceMetric, levels = c("accuracy","nMCC", "F1", "TPR", "TNR")), 
                        label = round(value, 4), 
                        fill = value)) +
    ggplot2::geom_tile(color = 'grey25') +
    scale_fill_gradientn(colors = hcl.colors(20, "RdYlGn"), values = c(0,0.5,1)) +
    ggplot2::ylab("Package") +
    ggplot2::xlab("Performance on Alexandrov et al. data split on TMB") +
    scale_x_discrete(position = "top") +
    theme(
        axis.text.x = element_text(angle = 0, ),
        text = ggplot2::element_text(size=10, family='Helvetica'),
        strip.background = ggplot2::element_blank(),
        panel.background = element_blank(),
        legend.title = element_blank()
    ) +
    ggplot2::geom_text(size = 2.9, angle = 0) +
    facet_grid(TMBcat~., labeller = labeller(TMBcat = TMB.labs)) 
```

```{r}
# facet labels
detectedKataegisInSample.labs <- c("Kataegis Samples", "Non Kataegis Samples")
names(detectedKataegisInSample.labs) <- c(TRUE, FALSE)

# Total runtimes of all PCF methods 
plotRunTimes.Alex <- resultsPerSampleAlex |> 
    ggplot2::ggplot(mapping = aes(x = factor(package, levels = arrange(performancePerSampleAlex, meanRuntime)$package), y = runTime/60, fill = package)) + 
    gghalves::geom_half_point_panel(size = 1, col = 'black', shape = 21) +
    gghalves::geom_half_boxplot(outlier.shape = NA) +
    ggplot2::scale_fill_manual(values = colors, guide = 'none') +
    ggplot2::scale_color_manual(values = colors, guide = 'none') +
    ggplot2::scale_y_continuous(trans = scales::pseudo_log_trans(), breaks = c(seq(0,1,by =.2), 0:20)) +
    ggplot2::labs(y = "Runtime in minutes<br>(Alexandrov et al.; log <sub>10</sub>)", x = 'Packages') +
    ggplot2::geom_text(data = resultsPerSampleAlex %>% dplyr::group_by(package) %>% dplyr::summarise(mean = mean(runTime, na.rm = T)), aes(x = package, y = -.1, label = round(mean, 1)), size = 3, fontface='bold') +
    theme_Job
```